Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-03-14 young_pp_discs.tex
done for now

\begin{abstract}
The formation and earliest stages of protoplanetary discs are still lacking observational constraints. ALMA will soon revolutionize this field, therefore it is important to provide predictions and help the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations based on radiative transfer modelling of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ dense core, with structure similar to the pre-stellar core L1544. We focus on simple species and conclude that  C$^{17}$O 3$\rightarrow$2, HCO$^+$ 3$\rightarrow$2, OCS 26$\rightarrow$25 and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ lines together allow for measurements of the disc structure and kinematics at all scales. 
\end{abstract}

@@

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). Although L1544 does not appear to contain a central protostar and disc itself, L1521F, a similar dense core, contains a central object which appears to be a low mass protostar and disc \citep{Bourke2006}, whilest having between half and three quarters of the mass of L1544. The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Figure \ref{fig:l1544_model} shows the physical parameters of the core model. \smallskip

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure used is from the same hydrodynamics simulation described in I2011, which is a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The star is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (figure \ref{rhoT}). For the non-LTE 3D radiative transfer modeling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y, and z and sample from this to create the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 1/100 is used throught the entirity of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history. In the dense conditions of the protoplanetary disc could be encouraging the dust grains to agglomerate and increase the size of the ice mantle through freeze out of volitiles. This could lead to changes in the dust opacity, however we have not attempted to include these effects in our model. The dust opacities used in the radiative transfer are not the same as the ones used in the hydrodynamic model (which used the dust opacities of \citealt{DAlessio2001}). However the results desciribed here do not depend on the details of the dust opacity and would be qualitatively similar if other opacities were used.\smallskip

@@

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
The radiative transfer program used is LIME (LIne Modeling Engine; \cite{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in section \ref{subsec:gridding}. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and unresolved turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation\footnote{For three dimensions, Delaunay trianglulating means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figure \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \citealt{Brinch2010}) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (Figure \ref{averages}).

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime2.eps}%Lime_grid3.eps
 \caption{A plot of the points selected by the gridding process and the paths down which photons can propagate for points in the central r,$\theta$ plane. The points are colour coded by the density distribution (in m$^{-3}$, as used in LIME) and are more concentrated at small radii and in the most dense regions. The circular high density region in the centre is the disc model and is 128$\,$au in diameter.}
 \label{grid}
\end{figure}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/continuum.eps}
 \caption{{\bf Left:}A 300$\,$GHz continuum image of the model, created from the average of ten LIME runs. {\bf Right:} the output of a single radiative transfer simulation, with artifacts due to finite gridding.}
 \label{averages}
\end{figure}

@@

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc). The inlination angle was varied from 15$^\circ$ to 75$^\circ$ to edge on, we will consider the 30$^\circ$ case as a 'typical' inclination with the 15$^\circ$ inclination discussed in section \ref{sec:alma_predictions}, the other inclinations simulated are shown in appendix \ref{sec:other_inc}. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity, intensity weighted velocity and channel maps (figures \ref{sim_all}, \ref{h2co_chanmap} and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope. This can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity (see section \ref{sec:alma_predictions}). Intensity weighted velocity maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

@@

Figure \ref{continuum} Shows the 1mm continuum emission and beam size in this configuration of ALMA. The spiral structure of the disc is clearly visible at this inclination and will be clearer the closer the disc is to being face on. We can get an estimate of the mass from a continuum image such as this using the equation $EQUATION$ \cite{equn_ref} which gives a mass of X showing that Y (continuum mass estimates are good/rubish for discs of this high mass/density).\smallskip

Figure \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$ line overlaid on the 1$\,$mm continuum emission. From this it can be seen that the absorption features trace the disc spiral structure whilst the line emission emanates from gaps in the outer disc, where there is less continuum to be absorbed. The central pannel is domminated by the envelope contribution, causing stron absorption through out the disc. The other pannels show the line following a rotation profile distorted by the presence of the spiral arms. \smallskip

Figure \ref{15deg} shows the C$^{17}$O 3$\rightarrow$2 integrated intensity and PV diagram for an inclination angle of 15$^\circ$ to edge on, showing that spiral structure can be seen in the position velocity diagram at low inclinations. The ``finger-like'' structures in the PV diagram are similar to those seen in the case of observations of the milky way and of simulations of spirals in the milky way (e.g. \citealt{Bissantz2003}). Whilest analysis of the detail of the spiral structure (number of arms, pitch angle etc.) would be difficult from such a PV diagram, detecting the presence of spiral waves, and therefore a potentially gravitationally unstable disc, is possible.\smallskip

@@

 \caption{CASA simulation of the C17O 3$\rightarrow$2 line with the disc inclided at 15$^\circ$ to edge on {\bf Top:} integrated intensity map. {\bf bottom:} position velocity diagram. In the position velocity diagram the spiral structures can be seen as narrow finger-like structures extending from the origin.}
 \label{15deg}
\end{figure}

@@

In this paper we have presented radiative transfer simulations of a hybrid model comprising a 0.39$\, M_\odot$ self gravitating disc with radius 64$\,$au showing spiral density waves, surrounded by an envelope simulated as a contracting 10$\,M_\odot$ BE-sphere. The main results of these simulations are as follows:\smallskip
$\bullet$ CASA simulations of this model show that at a distance of 100$\,$pc, extraction of kinematics and structural information from the continuum and molecular lines is possible with ALMA band 7.\\
$\bullet$ Our simulations show that molecular features are predominantly seen in absorption against the hot mid-plane of self-gravitating protoplanetary discs, with emission only being seen where voids in the structure of the disc do not provide a bright continuum source to be absorbed.\\
$\bullet$ The quiescent nature of the envelope around such discs only obscures lines within $\pm\,$0.5 km$\,$s$^{-1}$ of the systematic velocity.\\
$\bullet$ The lines studied (OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2) taken together allow all regions of the disc to be sampled and the rotation of the disc to be constrained from the inner edge to the outer edge.\\
$\bullet$ Spiral structure in young embedded discs are dectectable at a wide range of inclination angles, they can be resolved spatially at angles close to face on and can be infered from position velocity diagrams at low inclination angles.\smallskip

@@

Big up results here!!

@@

Simulations at inclinations other than 30$^\circ$ were also performed with C$^{17}$O. Figures B1-3 show the C$^{17}$O 3$\rightarrow$2 line in the same phisical model but at 15$^\circ$, 30$^\circ$, 45$^\circ$ and 75$^\circ$ to egde on. At 15$^\circ$ to edge on we can still see some structure in the integrated intensity map and the structure is also visible in the position velocity diagram. At larger incliations it gets more difficult to obtain any kinematic information but the physical structure of the disc becomes clearer.

>>>>

\begin{abstract}
The formation and earliest stages of protoplanetary discs are still lacking observational constraints. ALMA will soon revolutionise this field, therefore it is important to provide predictions and help the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations based on radiative transfer modelling of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ dense core, with structure similar to the pre-stellar core L1544. We focus on simple species and conclude that  C$^{17}$O 3$\rightarrow$2, HCO$^+$ 3$\rightarrow$2, OCS 26$\rightarrow$25 and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ lines together allow for measurements of the disc structure and kinematics at all scales. 
\end{abstract}

@@

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). Although L1544 does not appear to contain a central protostar and disc itself, L1521F, a similar dense core, contains a central object which appears to be a low mass protostar and disc \citep{Bourke2006}, whilst having between half and three quarters of the mass of L1544. The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Figure \ref{fig:l1544_model} shows the physical parameters of the core model. \smallskip

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure used is from the same hydrodynamics simulation described in I2011, which is a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The star is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (figure \ref{rhoT}). For the non-LTE 3D radiative transfer modelling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y, and z and sample from this to create the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 1/100 is used throughout the entirety of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history. In the dense conditions of the protoplanetary disc could be encouraging the dust grains to agglomerate and increase the size of the ice mantle through freeze out of volatiles. This could lead to changes in the dust opacity, however we have not attempted to include these effects in our model. The dust opacities used in the radiative transfer are not the same as the ones used in the hydrodynamic model (which used the dust opacities of \citealt{DAlessio2001}). However the results described here do not depend on the details of the dust opacity and would be qualitatively similar if other opacities were used.\smallskip

@@

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
The radiative transfer program used is LIME (LIne Modelling Engine; \cite{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in section \ref{subsec:gridding}. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and unresolved turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation\footnote{For three dimensions, Delaunay triangulation means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figure \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \citealt{Brinch2010}) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (Figure \ref{averages}).

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime2.eps}%Lime_grid3.eps
 \caption{A plot of the points selected by the griding process and the paths down which photons can propagate for points in the central r,$\theta$ plane. The points are colour coded by the density distribution (in m$^{-3}$, as used in LIME) and are more concentrated at small radii and in the most dense regions. The circular high density region in the centre is the disc model and is 128$\,$au in diameter.}
 \label{grid}
\end{figure}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/continuum.eps}
 \caption{{\bf Left:}A 300$\,$GHz continuum image of the model, created from the average of ten LIME runs. {\bf Right:} the output of a single radiative transfer simulation, with artifacts due to finite grid-ding.}
 \label{averages}
\end{figure}

@@

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc). The inclination angle was varied from 15$^\circ$ to 75$^\circ$ to edge on, we will consider the 30$^\circ$ case as a 'typical' inclination with the 15$^\circ$ inclination discussed in section \ref{sec:alma_predictions}, the other inclinations simulated are shown in appendix \ref{sec:other_inc}. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity, intensity weighted velocity and channel maps (figures \ref{sim_all}, \ref{h2co_chanmap} and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope. This can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity (see section \ref{sec:alma_predictions}). Intensity weighted velocity maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

@@

Figure \ref{continuum} Shows the 1mm continuum emission and beam size in this configuration of ALMA. The spiral structure of the disc is clearly visible at this inclination and will be clearer the closer the disc is to being face on. We can get an estimate of the mass from a continuum image such as this using the equation $EQUATION$ \cite{equn_ref} which gives a mass of X showing that Y (continuum mass estimates are good/rubbish for discs of this high mass/density).\smallskip

Figure \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$ line overlaid on the 1$\,$mm continuum emission. From this it can be seen that the absorption features trace the disc spiral structure whilst the line emission emanates from gaps in the outer disc, where there is less continuum to be absorbed. The central panel is dominated by the envelope contribution, causing Strong absorption through out the disc. The other panels show the line following a rotation profile distorted by the presence of the spiral arms. \smallskip

Figure \ref{15deg} shows the C$^{17}$O 3$\rightarrow$2 integrated intensity and PV diagram for an inclination angle of 15$^\circ$ to edge on, showing that spiral structure can be seen in the position velocity diagram at low inclinations. The ``finger-like'' structures in the PV diagram are similar to those seen in the case of observations of the milky way and of simulations of spirals in the milky way (e.g. \citealt{Bissantz2003}). Whilst analysis of the detail of the spiral structure (number of arms, pitch angle etc.) would be difficult from such a PV diagram, detecting the presence of spiral waves, and therefore a potentially gravitationally unstable disc, is possible.\smallskip

@@

 \caption{CASA simulation of the C17O 3$\rightarrow$2 line with the disc inclined at 15$^\circ$ to edge on {\bf Top:} integrated intensity map. {\bf bottom:} position velocity diagram. In the position velocity diagram the spiral structures can be seen as narrow finger-like structures extending from the origin.}
 \label{15deg}
\end{figure}

@@

In this paper we have presented radiative transfer simulations of a hybrid model comprising a 0.39$\, M_\odot$ self gravitating disc with radius 64$\,$au showing spiral density waves, surrounded by an envelope simulated as a contracting 10$\,M_\odot$ BE-sphere. The main results of these simulations are as follows:\newline
$\bullet$ CASA simulations of this model show that at a distance of 100$\,$pc, extraction of kinematics and structural information from the continuum and molecular lines is possible with ALMA band 7.\\
$\bullet$ Our simulations show that molecular features are predominantly seen in absorption against the hot mid-plane of self-gravitating protoplanetary discs, with emission only being seen where voids in the structure of the disc do not provide a bright continuum source to be absorbed.\\
$\bullet$ The quiescent nature of the envelope around such discs only obscures lines within $\pm\,$0.5 km$\,$s$^{-1}$ of the systematic velocity.\\
$\bullet$ The lines studied (OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2) taken together allow all regions of the disc to be sampled and the rotation of the disc to be constrained from the inner edge to the outer edge.\\
$\bullet$ Spiral structure in young embedded discs are detectable at a wide range of inclination angles, they can be resolved spatially at angles close to face on and can be inferred from position velocity diagrams at low inclination angles.\smallskip

@@

These results demonstrate that with next generation millimetre interferometers we will be able to probe the very earliest stages of circumstellar discs which feed their growing protostar and eventually could become a planetary system. Our results show that these types of observations would be able to detect whether or not these young disc are gravitationally unstable. If such gravitationally instabilities exist they would have significant implications on the evolution of the disc, such as fragmentation and planetesimal formation \citep{Boley2009,Johnson2013,Gibbons2012} and chemical evolution I2011.

@@

Simulations at inclinations other than 30$^\circ$ were also performed with C$^{17}$O. Figures B1-3 show the C$^{17}$O 3$\rightarrow$2 line in the same physical model but at 15$^\circ$, 30$^\circ$, 45$^\circ$ and 75$^\circ$ to edge on. At 15$^\circ$ to edge on we can still see some structure in the integrated intensity map and the structure is also visible in the position velocity diagram. At larger inclinations it gets more difficult to obtain any kinematic information but the physical structure of the disc becomes clearer.
